---
title: "Cyclistic 2022 Case Study"
author: "Leopoldine Mirtil"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


### Disclaimer

This analysis was made from the _Cyclistic Bike Share Case Study: How Does a Bike-Share Navigate Speedy Success?_ offered through the Google Data Analytics Certificate program on Coursera.com. The data has been made publicly available by the Motivate International Inc license. The data was originally downloaded from the 2022 monthly [Divvy trip dataset](https://divvy-tripdata.s3.amazonaws.com/index.html). 


## Introduction

### Scenario

You are a junior data analyst working in the marketing analyst team at Cyclistic, a bike-share company in Chicago. The director of marketing believes the company’s future success depends on maximizing the number of annual memberships. Therefore, your team wants to understand how casual riders and annual members use Cyclistic bikes differently. From these insights, your team will design a new marketing strategy to convert casual riders into annual members. But first, Cyclistic executives must approve your recommendations, so they must be backed up with compelling data insights and professional data visualizations.

### Task
1. How do annual members and casual riders use Cyclistic bikes differently?
2. Why would casual riders buy Cyclistic annual memberships?
3. How can Cyclistic use digital media to influence casual riders to become members?


## Let's Get to Work

### Step 1 - Import Data

#### Load Packages

```{r, message=FALSE}
#for data manipulation
library(dplyr)
library(lubridate)
library(tidyr)
library(tidyverse)

#for visualizations
library(ggplot2)

#for documentation/markdown
library(knitr)
```



#### Import Data Files

```{r, message=FALSE}
setwd("C:/Users/Leopoldine/Desktop/Mine/Coding Projects & Portfolio/Cyclistic Case Study/Cyclistic2022/00_raw_data")

m1_2022 <- read.csv("202201-divvy-tripdata.csv")
m2_2022 <- read.csv("202202-divvy-tripdata.csv")
m3_2022 <- read.csv("202203-divvy-tripdata.csv")
m4_2022 <- read.csv("202204-divvy-tripdata.csv")
m5_2022 <- read.csv("202205-divvy-tripdata.csv")
m6_2022 <- read.csv("202206-divvy-tripdata.csv")
m7_2022 <- read.csv("202207-divvy-tripdata.csv")
m8_2022 <- read.csv("202208-divvy-tripdata.csv")
m9_2022 <- read.csv("202209-divvy-publictripdata.csv")
m10_2022 <- read.csv("202210-divvy-tripdata.csv")
m11_2022 <- read.csv("202211-divvy-tripdata.csv")
m12_2022 <- read.csv("202212-divvy-tripdata.csv")
```



### Step 2 - Merge Data Sets 

```{r}
#combine all files
total_trips <- bind_rows(m1_2022, m2_2022, m3_2022, m4_2022, m5_2022, m6_2022, m7_2022, m8_2022, m9_2022, m10_2022, m11_2022, m12_2022)

#inspect new data frame
str(total_trips)  
```



### Step 3 - Add New Columns

```{r}
#get initial date riders got the bikes and for later use 
total_trips$date <- as.Date(total_trips$started_at) 

#month of start date to identify patterns and trends
total_trips$month <- format(as.Date(total_trips$date), "%m")  

#full weekday name of the date(Sunday, Saturday...)
total_trips$day_of_week <- format(as.Date(total_trips$date), "%A")  

#rental duration in seconds
total_trips$duration <- difftime(total_trips$ended_at,total_trips$started_at) 
```


### Step 4 - Cleaning Process 

#### Inspect Modified Data Frame

```{r}
str(total_trips) 
```


#### Change Data Type of Column

```{r}
#copy data before further modifications
total_tripsv2 <- total_trips

#convert to numeric data type & remove 'secs' unit
total_tripsv2$duration <- as.numeric(total_tripsv2$duration)

# confirm data type change
str(total_tripsv2$duration)
```


#### Remove Negative Values 

```{r}
total_tripsv2 <- total_tripsv2[!(total_tripsv2$duration<0),]
```


#### Remove Unneeded Columns

```{r}
#remove 6 columns: both station Id, lat & long 
total_tripsv2 <- total_tripsv2[-c(6, 8:12)] 
```


#### Rename Columns

```{r}
total_tripsv2 <- rename(total_tripsv2, bike_type=rideable_type, rider_type=member_casual)
```


#### Check for Missing & Blank Values

```{r}
#location & count of any missing/NA values
sapply(total_tripsv2, function(x) sum(is.na(x)))
```

```{r}
#location & count of blank cells 
sapply(total_tripsv2, function(x) sum(x==""))
```


#### Remove Empty Rows 

```{r}
total_tripsv2 <- total_tripsv2[!(total_tripsv2$start_station_name %in% ""),]
total_tripsv2 <- total_tripsv2[!(total_tripsv2$end_station_name %in% ""),]
```


```{r}
#confirm no more empty rows
sapply(total_tripsv2, function(x) sum(x==""))
```
#### Inspect Final, Modified Data Frame

```{r}
str(total_tripsv2) 
```


```{r, results='hide', echo=FALSE, message=FALSE}
#export final modified data frame
setwd("C:/Users/Leopoldine/Desktop/Mine/Coding Projects & Portfolio/Cyclistic Case Study/Cyclistic2022/01_tidy_data")

#final export
write.csv(total_tripsv2, "total_tripsv2.csv", row.names = FALSE)
```


## Descriptive Analysis

#### Summary Duration by Rider Type 

```{r, message=FALSE}
total_tripsv2 %>%                            
  group_by(rider_type) %>%
    summarize(min = min(duration), 
    q1 = quantile(duration, 0.25), 
    median = median(duration), 
    mean = sprintf("%.2f", mean(duration)), 
    q3 = quantile(duration, 0.75), 
    max = max(duration)) 
```


#### Summary Duration by Rider & Bike Types 

```{r, message=FALSE}
total_tripsv2 %>%                            
  group_by(rider_type, bike_type) %>%
    summarize(min = min(duration), 
    q1 = quantile(duration, 0.25), 
    median = median(duration), 
    mean = sprintf("%.2f", mean(duration)),  
    q3 = quantile(duration, 0.75), 
    max = max(duration))
```


#### Total Trip Count by Rider Type

```{r}
total_tripsv2 %>% count(rider_type, sort = TRUE)
```


#### Total Trip Count by Bike Type

```{r}
total_tripsv2 %>% count(bike_type, sort = TRUE)
```


#### Total Trip Count by Bike & Rider Types

```{r}
total_tripsv2 %>% count(bike_type, rider_type, sort = TRUE)
```


#### Total Weekday Ride Count by Rider Type 

```{r, message=FALSE}
total_tripsv2$day_of_week <- ordered(total_tripsv2$day_of_week, levels=c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday"))
total_tripsv2 %>%                            
  group_by(rider_type, day_of_week) %>%
    summarize(number_of_rides = n()) %>%
    arrange(day_of_week, rider_type) 
```


#### Average Weekday Ride Duration by Rider Type

```{r, message=FALSE}
total_tripsv2$day_of_week <- ordered(total_tripsv2$day_of_week, levels=c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday"))
total_tripsv2 %>%                            
  group_by(rider_type, day_of_week) %>%
    summarise(average_duration = sprintf("%.2f", mean(duration))) %>%
    arrange(day_of_week, rider_type) 
```


#### Total Monthly Number of Rides by Rider Type

```{r, message=FALSE}
total_tripsv2 %>%                            
  group_by(rider_type, month) %>%
    summarise(number_of_rides = n()) %>%
    arrange(month, rider_type) %>%
    print(n=24)
```


#### Top 5 Destinations by Riders

```{r, message=FALSE, error=FALSE}
total_tripsv2 %>%
  group_by(rider_type, end_station_name) %>%
    summarize(number_of_rides = n()) %>%
    arrange(rider_type, desc(number_of_rides)) %>%
    slice(1:5) 
``` 


## Plots and Visualizations

#### Preferred Bikes among Riders

```{r, echo=FALSE}
total_tripsv2$bike_type <-factor(total_tripsv2$bike_type,(levels =c("docked_bike", "classic_bike", "electric_bike"))) 
total_tripsv2 %>% 
  count(bike_type, rider_type, sort = TRUE) %>%
  group_by(rider_type) %>%
  mutate(pct= prop.table(n) * 100) %>%
    arrange(bike_type) %>%
      ggplot(aes(fill=bike_type, y=pct, x=rider_type)) +
        geom_bar(stat = "identity") +
        ylab("Percentage of Rides") +
        theme(axis.title.x=element_blank()) +
          geom_text(aes(label=paste0(sprintf("%1.2f", pct),"%")),
            position=position_stack(vjust=0.5), color = "white") + 
            guides(fill=guide_legend(title="Bike Type")) 
```


The classic bike is the most preferred bike with over half of member riders and about fifty percent of casual riders. The docked bike was only used by casual riders, but made up a small percentage of their total ride count.


#### Total Weekday Ride Count

```{r, echo=FALSE, message=FALSE}
options(scipen=999) #remove scientific notation from y-axis
library(scales) # load pkg to scale y-axis

total_tripsv2 %>% 
  mutate(weekday = wday(started_at, label=TRUE)) %>%
  group_by(rider_type, weekday) %>%
    summarize(number_of_rides = n()) %>%
    arrange(rider_type, weekday) %>%
ggplot(mapping = aes(x=weekday, y=number_of_rides, fill=rider_type)) + 
  geom_col(position = "dodge") +
  ylab("Ride Count") +
  theme(axis.title.x=element_blank()) +
  scale_y_continuous(labels = scales::comma) + 
  guides(fill=guide_legend(title="Rider Type")) 
```

Annual members consistently had a larger number of rides Monday to Friday, with the maximum number of rides on occurring on Thursday. On the weekends, casual riders had a higher number of riders than members, with the highest amount occurring on Saturday.


#### Average Weekday Trip Duration

```{r, echo=FALSE, message=FALSE}
total_tripsv2 %>% 
  mutate(weekday = wday(started_at, label=TRUE)) %>%
  group_by(rider_type, weekday) %>%
    summarise(average_duration = mean(duration)) %>%
    arrange(rider_type, weekday) %>%
ggplot(mapping = aes(x=weekday, y=average_duration, fill=rider_type)) + 
  geom_col(position = "dodge") + 
  ylab("Average Trip Duration(in seconds)") +
  theme(axis.title.x=element_blank()) +
  scale_y_continuous(labels = scales::comma) +
  guides(fill=guide_legend(title="Rider Type"))
```


The average ride duration during the week among casual riders was consistently twice the length of member riders during the week. The average duration among both rider types is highest on the weekends. Meanwhile, the average ride duration among annual members was steadily under 1,000 seconds throughout the week. 


#### Total Monthly Ride Count 

```{r, echo=FALSE, message=FALSE}
total_tripsv2 %>% 
  mutate(months = month(started_at, label=TRUE)) %>%
  group_by(rider_type, months) %>%
    summarise(number_of_rides = n()) %>%
      arrange(rider_type, months) %>%
ggplot(aes(x = months, y = number_of_rides, color = rider_type, group = rider_type)) +
    geom_line(linewidth = 1.5) +
    geom_point(size = 3) + 
    ylab("Ride Numbers") +
    theme(axis.title.x=element_blank()) +
    scale_y_continuous(labels = scales::comma) +
    guides(color=guide_legend(title="Rider Type")) 
```


Annual members consistently had a higher number of rides across the year when compared to casual riders. The maximum number of rides occurred in July for casual riders and in August for members. There is a noticeable drop in the number of rides after July among casual riders and after August among annual members. This corresponds with the upcoming fall and winter seasons. The maximum amount of rides for both rider types occurs during the spring and summer seasons. 


#### Top 5 End Station Destinations

```{r, echo=FALSE, message=FALSE}
total_tripsv2 %>%
  group_by(rider_type, end_station_name) %>%
    summarize(number_of_rides = n()) %>%
    arrange(rider_type, desc(number_of_rides)) %>%
    slice(1:5) %>%
  ggplot(aes(x=rider_type, y=number_of_rides, fill=end_station_name, label=number_of_rides)) + 
      geom_bar(stat = "identity") + 
      geom_text(aes(label=format(number_of_rides, big.mark = ",")), position=position_stack(vjust=0.5), color="white") +
      scale_y_continuous(labels = scales::comma) +
    ylab("Ride Numbers") +
    theme(axis.title.x=element_blank()) +
    guides(fill=guide_legend(title="End Station"))
```


The top 5 destinations (end stations) among both riders are different locations. The Streeter Dr & Grand Ave station is the top destination among casual riders while for member riders it is the Kingsbury St & Kinzie St station. The numbers are more consistent among members than casual riders.

### Recommendations


**Increase Classic Bikes**

Replace docked bikes with new classic bikes to bring in more casual riders. Since the classic bike is the most popular among casual riders, increasing the amount available at stations will help bring in more casual riders and convert them to members. This will also have the added benefit of increasing the available bikes to members as well.


**Membership Specials**

Offer new membership specials that would offer membership to casual riders at a discount. Since there is a marked increase in the number of casual riders on the weekend and approaching the summer from May to July, it would be  advantageous to offer Weekend, Spring and Summer Specials. The company can also show highlight perks of membership through digital media to influence casual riders to become members. This can include posting member testimonials and ads highlighting the benefits of the bike rides (save time or money, enjoy the sites, get in shape, etc.) 


**Advertise Top 5 Destinations**

Use digital media and ads near start stations to advertise the top 5 end stations used by riders. Since the top 5 destinations of casual riders differ from those of members, the advertisements will help attract more casual riders and convert them to membership. 


